---
title: "Rideship data analysis"
pdf_document: default
---
The goal from this analysis is to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, we will design a new marketing strategy to convert casual riders into annual members. 
For the analysis we used the past 12 months worth of data from the Cyclistic database.
All data can be found on <https://divvy-tripdata.s3.amazonaws.com/index.html>

# Setup

## Loading the necessary packages

```{r package loading, include=FALSE}

library(readr)
library(tidyverse) #for data import and wrangling
library(dplyr)
library(lubridate) #for date functions
library(ggplot2) #for visualizations

```

## Importing the .csv files

We download all the data files from <https://divvy-tripdata.s3.amazonaws.com/index.html> then we import them in our script

```{r file imports, include=FALSE}

dataset1 <- read_csv("citibike data/202107-divvy-tripdata.csv")
dataset2 <- read_csv("citibike data/202108-divvy-tripdata.csv")
dataset3 <- read_csv("citibike data/202109-divvy-tripdata.csv")
dataset4 <- read_csv("citibike data/202110-divvy-tripdata.csv")
dataset5 <- read_csv("citibike data/202111-divvy-tripdata.csv")
dataset6 <- read_csv("citibike data/202112-divvy-tripdata.csv")
dataset7 <- read_csv("citibike data/202201-divvy-tripdata.csv")
dataset8 <- read_csv("citibike data/202202-divvy-tripdata.csv")
dataset9 <- read_csv("citibike data/202203-divvy-tripdata.csv")
dataset10 <- read_csv("citibike data/202204-divvy-tripdata.csv")
dataset11 <- read_csv("citibike data/202205-divvy-tripdata.csv")
dataset12 <- read_csv("citibike data/202206-divvy-tripdata.csv")

```

## Merging the files into one dataset

We merge all the data files into one Dataset for easier analysis, then we print the resulting number of rows

```{r merging datasets}

merged_data <- rbind(dataset1,dataset2,dataset3,dataset4,dataset5,dataset6,dataset7,dataset8,dataset9,dataset10,dataset11,dataset12)

nrows <- nrow(merged_data)
print(paste("Number of rows:",nrows))

```

# Data Cleaning
## Removing duplicates and formatting date/time

To organize and clean our data we will format the dates and times into a consistent format, as well as delete the duplicate rows when found

```{r removing duplicates, results='hide'}

data_no_dups <- merged_data[!duplicated(merged_data$ride_id),]

data_no_dups$started_at <- as.POSIXct(data_no_dups$started_at, "%Y-%m-%d %H:%M:%S")
data_no_dups$ended_at <- as.POSIXct(data_no_dups$ended_at, "%Y-%m-%d %H:%M:%S")

```

## Removing missing values

We figure out the number of missing values in each column. 

```{r missing values per column}
colSums(is.na(data_no_dups))
```

We find out that only the columns end_lat and end_lng contain missing values.

```{r removing missing valies}

#remove duplicate values
dataset_no_na <- drop_na(data_no_dups, start_lat, end_lat, start_lng, end_lng)

deleted_rows <- nrow(merged_data) - nrow(dataset_no_na)
print(paste("Deleted rows:",deleted_rows))

```

## Creating additional columns for aggregation

We extract additional columns from the already-existing date columns. We get days, months, and years out for us to have more flexibility in future aggreagations.

```{r extracting date columns, results='hide'}

dataset_no_na$start_date <- as.Date(dataset_no_na$started_at)
dataset_no_na$start_month <- format(as.Date(dataset_no_na$started_at),"%m")
dataset_no_na$start_day <- format(as.Date(dataset_no_na$started_at),"%d")
dataset_no_na$start_hour <- format(as.POSIXct(dataset_no_na$started_at), format = "%H")
dataset_no_na$start_year <- format(as.Date(dataset_no_na$started_at),"%Y")
dataset_no_na$start_day_of_week <- format(as.Date(dataset_no_na$started_at),"%A")

dataset_no_na$end_date <- as.Date(dataset_no_na$ended_at)
dataset_no_na$end_month <- format(as.Date(dataset_no_na$ended_at),"%m")
dataset_no_na$end_day <- format(as.Date(dataset_no_na$ended_at),"%d")
dataset_no_na$end_year <- format(as.Date(dataset_no_na$ended_at),"%Y")
dataset_no_na$end_day_of_week <- format(as.Date(dataset_no_na$ended_at),"%A")

```


## Calculating ride lengths

We calculate the ride lengths for each entry in minutes
```{r calculating ride lengths, results='hide'}

# add a ride length column
dataset_no_na <- mutate(dataset_no_na, ride_length = (ended_at-started_at))
#converting it into numeric and dividing by 60 for minutes
dataset_no_na$ride_length <- as.numeric(as.character(dataset_no_na$ride_length))
dataset_no_na = mutate(dataset_no_na, ride_length = round(ride_length/60,digits=2))

```

We take a peek at our data

```{r aggregating data}

aggregate(dataset_no_na$ride_length ~dataset_no_na$member_casual, FUN =mean)
aggregate(dataset_no_na$ride_length ~dataset_no_na$member_casual, FUN =median)
aggregate(dataset_no_na$ride_length ~dataset_no_na$member_casual, FUN =max)
aggregate(dataset_no_na$ride_length ~dataset_no_na$member_casual, FUN =min)

```
We notice there are negative values in this columns, we proceed to clean up the data by deleting those entries.

```{r deleting negative values, results='hide'}

#removing negative ride lengths:
clean_dataset <- dataset_no_na[!(dataset_no_na$ride_length <=0),]

```

# Descriptive Analysis
## Comparison across ride types

```{r summarizing data}

#summarizing data
summary <- clean_dataset %>% 
  summarize(mean(ride_length),median(ride_length),max(ride_length),min(ride_length))

print(summary)

```


```{r comparing between rid types}

#displaying the days of the week in order
clean_dataset$start_day_of_week <- ordered(clean_dataset$start_day_of_week, levels = c("Saturday","Sunday","Monday","Tuesday","Wednesday","Thursday","Friday"))

#analyzing ridership data by type and weekday
clean_dataset %>% 
  mutate(weekday = wday(started_at, label=TRUE)) %>%
  group_by(member_casual,weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual,weekday)

```
# Visualizations
## Visualizing the number of rides by rider type

```{r number of rides/rider type}

clean_dataset %>%
  mutate(weekday=wday(started_at, label=TRUE)) %>%
  group_by(member_casual,weekday) %>%
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>%
  arrange(member_casual,weekday) %>%
  ggplot(aes(x=weekday, y=number_of_rides, fill=member_casual)) +
  geom_col(position="dodge")

```

## Visualizing the average duration by rider type

```{r average duration/rider type}

clean_dataset %>%
  mutate(weekday =wday(started_at,label=TRUE)) %>%
  group_by(member_casual,weekday) %>%
  summarise(number_of_rides=n(),average_duration=mean(ride_length)) %>%
  arrange(member_casual,weekday) %>%
  ggplot(aes(x=weekday,y=average_duration, fill=member_casual)) +
  geom_col(position="dodge")

```

## Visualising number of rides per month

Key insights: the busiest months during the year are from May to September, mainly dominated by casual riders

```{r}

clean_dataset %>%
  group_by(member_casual,start_month) %>%
  summarise(number_of_rides=n()) %>%
  arrange(member_casual, start_month) %>%
  ggplot(aes(x=start_month, y=number_of_rides, fill = member_casual)) +
  geom_col(position="dodge")

```

## Visualizing busiest hours during the day

Key insights: the busiest hours during the day are during the evenings also dominated by casual riders

```{r}

clean_dataset %>%
  group_by(member_casual,start_day,start_hour) %>%
  summarise(number_of_rides=n()) %>%
  arrange(member_casual, start_day, start_hour) %>%
  ggplot(aes(x=start_hour, y=number_of_rides, fill = member_casual)) +
  geom_bar(position="stack", stat="identity")

```

# Recommendations
Here are the top 3 recommendations to convert casual riders into members:

 1. Discount memberships for summer holidays and weekends
 2. Special points gained during evening hours for members + potential for doubling point gain at certain hours
 3. A one time payment prior to every casual riding period + usual hourly rates